---
title: "Cyclistic Programme Monitor"
subtitle: "Chicago Divvy · 2021–2025"
format:
  dashboard:
    theme: [flatly, styles.scss]
    orientation: rows
    scrolling: true
    embed-resources: true
    nav-buttons:
      - icon: journal-text
        href: https://github.com/ThacienneUwimanayantumye/Cyclistic-Bike-Share-Analysis/blob/main/docs/METHODS.md
        aria-label: Methods
      - icon: github
        href: https://github.com/ThacienneUwimanayantumye/Cyclistic-Bike-Share-Analysis
        aria-label: Source
knitr:
  opts_chunk:
    fig-width: 8.8
    fig-height: 4.5
    fig-dpi: 120
    out-width: 100%
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(ggplot2)
  library(tidyr)
  library(scales)
})

derived <- file.path("..", "data", "derived")
if (!file.exists(file.path(derived, "member_share.csv"))) {
  derived <- file.path("data", "derived")
}

read_derived <- function(name) {
  read_csv(file.path(derived, paste0(name, ".csv")), show_col_types = FALSE)
}

member_share <- read_derived("member_share")
programme_month <- read_derived("programme_month")
quality_month <- read_derived("quality_month")
timing <- read_derived("timing")
duration_month <- read_derived("duration_month")
duration_bins <- read_derived("duration_bins")
bike_type <- read_derived("bike_type")
rider_profile <- read_derived("rider_profile")

pal <- c(Member = "#215C8C", Casual = "#C05621")

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      legend.position = "top",
      legend.title = element_blank(),
      legend.margin = margin(0, 0, 0, 0),
      legend.box.margin = margin(0, 0, -6, 0),
      panel.grid.minor = element_blank(),
      plot.margin = margin(6, 18, 6, 6),
      plot.subtitle = element_blank(),
      plot.caption = element_text(colour = "grey45", hjust = 0, size = 9),
      axis.title = element_text(colour = "grey30", size = rel(0.9)),
      axis.title.y = element_text(margin = margin(r = 6))
    )
)

as_programme <- function(x) {
  factor(ifelse(x == "member", "Member", "Casual"), levels = c("Member", "Casual"))
}

to_date <- function(ym) as.Date(paste0(ym, "-01"))
month_num <- function(ym) as.integer(substr(ym, 6, 7))

for (nm in c("member_share", "programme_month", "duration_month", "quality_month")) {
  assign(nm, mutate(get(nm), date = to_date(year_month)))
}
programme_month$programme <- as_programme(programme_month$member_casual)
duration_month$programme <- as_programme(duration_month$member_casual)

date_axis <- scale_x_date(
  date_breaks = "1 year", date_labels = "%Y",
  expand = expansion(mult = c(0.01, 0.01))
)

n_trips <- sum(member_share$n_trips)
n_casual <- sum(member_share$n_casual)
n_member <- sum(member_share$n_member)
share_casual_all <- n_casual / n_trips
share_member_all <- n_member / n_trips
n_in <- sum(quality_month$n_in)
drop_rate <- 1 - n_trips / n_in

share_by_programme <- function(data, condition) {
  data |>
    mutate(programme = as_programme(member_casual)) |>
    group_by(programme) |>
    summarise(value = sum(n_trips[{{ condition }}]) / sum(n_trips), .groups = "drop") |>
    tibble::deframe()
}

weekend <- share_by_programme(timing, weekday %in% c("Saturday", "Sunday"))
peak <- share_by_programme(
  rider_profile,
  day_type == "Weekday" & time_block %in% c("AM peak", "PM peak")
)
long_trip <- share_by_programme(
  duration_bins,
  !duration_bin %in% c("1-5", "5-10", "10-15", "15-20", "20-30")
)

med_duration <- duration_month |>
  group_by(programme) |>
  summarise(med = weighted.mean(duration_median_min, n_trips), .groups = "drop") |>
  tibble::deframe()

commuter_like <- rider_profile |>
  filter(
    day_type == "Weekday",
    time_block %in% c("AM peak", "PM peak"),
    duration_band == "Under 15 min"
  ) |>
  group_by(member_casual) |>
  summarise(n = sum(n_trips), .groups = "drop")
casual_commuter_n <- commuter_like$n[commuter_like$member_casual == "casual"]
casual_commuter_share <- casual_commuter_n /
  sum(rider_profile$n_trips[rider_profile$member_casual == "casual"])

helpers_file <- file.path("..", "dashboard", "helpers.R")
if (!file.exists(helpers_file)) helpers_file <- file.path("dashboard", "helpers.R")
sys.source(helpers_file, envir = environment())
```

# The difference

## {.flow}

::: {.hero-finding}
<p class="kicker">Chicago Divvy · 2021–2025</p>
<h2>Members commute; casual riders leisure-ride.</h2>
<p>
Of `r comma(n_trips, scale = 1e-6, accuracy = 0.1, suffix = " million")` cleaned trips, casual riders take `r percent(share_casual_all, accuracy = 1)` and members `r percent(share_member_all, accuracy = 1)`. Weekend share `r percent(weekend[["Casual"]], accuracy = 1)` vs `r percent(weekend[["Member"]], accuracy = 1)`; median `r number(med_duration[["Casual"]], accuracy = 0.1)` vs `r number(med_duration[["Member"]], accuracy = 0.1)` min. <strong>`r comma(casual_commuter_n, scale = 1e-6, accuracy = 0.1, suffix = " million")` casual trips</strong> are already short weekday peak-hour rides — the conversion target.
</p>
<p class="caveat">Trip-level only: Divvy files have no rider id.</p>
:::

## {.flow}

```{r}
#| results: asis
#| echo: false
#| fill: false
cat(kpi_strip_html(
  list(
    n = n_trips,
    n_member = n_member,
    n_casual = n_casual,
    share_member = share_member_all,
    share_casual = share_casual_all
  ),
  weekend,
  med_duration,
  casual_commuter_n,
  casual_commuter_share
))
```

## Signature {.flow}

### {width="56%"}

```{r}
#| title: "Members peak twice on weekdays; casual riders peak once in the afternoon"
#| fill: false
#| fig-width: 8.6
#| fig-height: 4.6
hourly <- timing |>
  filter(!weekday %in% c("Saturday", "Sunday")) |>
  mutate(programme = as_programme(member_casual)) |>
  group_by(programme, hour) |>
  summarise(n_trips = sum(n_trips), .groups = "drop_last") |>
  mutate(share = n_trips / sum(n_trips)) |>
  ungroup()

ggplot(hourly, aes(hour, share, colour = programme)) +
  annotate("rect", xmin = 6, xmax = 9, ymin = -Inf, ymax = Inf, fill = "grey88", alpha = 0.55) +
  annotate("rect", xmin = 16, xmax = 19, ymin = -Inf, ymax = Inf, fill = "grey88", alpha = 0.55) +
  geom_line(linewidth = 1.15) +
  scale_colour_manual(values = pal) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.04))) +
  scale_x_continuous(breaks = seq(0, 23, 3), labels = function(h) sprintf("%02d:00", h)) +
  labs(
    x = "Hour the trip started", y = "Share of weekday trips",
    caption = "Shaded bands: 06–09 and 16–19."
  )
```

### {width="44%"}

```{r}
#| title: "Casual riding is weekend and long; member riding is weekday and short"
#| fill: false
#| fig-width: 7.2
#| fig-height: 4.6
signature <- tibble::tibble(
  metric = factor(
    c("Weekend", "Weekday commute window", "Longer than 30 minutes"),
    levels = c("Longer than 30 minutes", "Weekday commute window", "Weekend")
  ) |> rep(each = 2),
  programme = factor(rep(c("Member", "Casual"), 3), levels = c("Member", "Casual")),
  value = c(
    weekend[["Member"]], weekend[["Casual"]],
    peak[["Member"]], peak[["Casual"]],
    long_trip[["Member"]], long_trip[["Casual"]]
  )
)

ggplot(signature, aes(value, metric, fill = programme)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.64) +
  geom_text(
    aes(label = percent(value, accuracy = 1)),
    position = position_dodge(width = 0.72),
    hjust = -0.12, size = 3.5, colour = "grey20"
  ) +
  scale_fill_manual(values = pal) +
  scale_x_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.22))
  ) +
  labs(x = "Share of the group's trips", y = NULL)
```

## Opportunity {.flow}

### {width="56%"}

```{r}
#| title: "Commute-like casual trips show up in summer, not winter"
#| fill: false
#| fig-width: 8.6
#| fig-height: 4.4
commute_pool <- rider_profile |>
  filter(
    day_type == "Weekday",
    time_block %in% c("AM peak", "PM peak"),
    duration_band == "Under 15 min"
  ) |>
  mutate(
    programme = as_programme(member_casual),
    month = factor(
      month_num(year_month),
      levels = 1:12,
      labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    )
  ) |>
  group_by(programme, month) |>
  summarise(n_trips = sum(n_trips), .groups = "drop")

ggplot(commute_pool, aes(month, n_trips, fill = programme)) +
  annotate("rect", xmin = 4.5, xmax = 9.5, ymin = -Inf, ymax = Inf, fill = "grey88", alpha = 0.55) +
  geom_col(position = position_dodge(width = 0.78), width = 0.72) +
  scale_fill_manual(values = pal) +
  scale_y_continuous(
    labels = label_number(scale = 1e-3, suffix = "k", accuracy = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = NULL, y = "Commute-like trips",
    caption = paste0(
      "Commute-like = weekday, 06–09 or 16–19, under 15 min. ",
      comma(casual_commuter_n, scale = 1e-6, accuracy = 0.1, suffix = " million"),
      " casual trips in this segment (",
      percent(casual_commuter_share, accuracy = 0.1),
      " of casual riding). Months pooled across 2021–2025."
    )
  )
```

### {width="44%"}

```{r}
#| title: "Member trips bunch under 15 minutes; casual trips have a long tail"
#| fill: false
#| fig-width: 7.2
#| fig-height: 4.4
dur_dist <- duration_bins |>
  mutate(
    programme = as_programme(member_casual),
    duration_bin = recode(
      as.character(duration_bin),
      "1-5" = "1–10", "5-10" = "1–10",
      "10-15" = "10–20", "15-20" = "10–20",
      "20-30" = "20–30",
      "30-45" = "30–45",
      "45-60" = "45–60",
      "60-90" = "60+", "90-120" = "60+", "120+" = "60+"
    ),
    duration_bin = factor(
      duration_bin,
      levels = c("1–10", "10–20", "20–30", "30–45", "45–60", "60+")
    )
  ) |>
  group_by(programme, duration_bin) |>
  summarise(n_trips = sum(n_trips), .groups = "drop_last") |>
  mutate(share = n_trips / sum(n_trips)) |>
  ungroup()

ggplot(dur_dist, aes(duration_bin, share, fill = programme)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.72) +
  scale_fill_manual(values = pal) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.04))) +
  labs(x = "Trip length (minutes)", y = "Share of trips")
```

# Week and season

## Heatmaps {.flow}

### Members concentrate on weekday evenings; casual riders spread into weekends

```{r}
#| title: "Members concentrate on weekday evenings; casual riders spread into weekends"
#| fill: false
#| fig-width: 12.2
#| fig-height: 5.4
heat <- timing |>
  mutate(programme = as_programme(member_casual)) |>
  group_by(programme, weekday, hour) |>
  summarise(n_trips = sum(n_trips), .groups = "drop_last") |>
  mutate(share = n_trips / sum(n_trips)) |>
  ungroup() |>
  mutate(weekday = factor(weekday, levels = rev(c(
    "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"
  ))))

ggplot(heat, aes(hour, weekday, fill = share)) +
  geom_tile() +
  facet_wrap(~programme) +
  scale_fill_viridis_c(
    option = "mako", direction = -1, labels = percent_format(accuracy = 0.1),
    name = "Share of the group's trips",
    guide = guide_colourbar(barwidth = 14, barheight = 0.55, title.position = "top")
  ) +
  scale_x_continuous(
    breaks = seq(0, 21, 3),
    labels = function(h) sprintf("%02d", h),
    expand = c(0, 0)
  ) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(x = "Hour the trip started", y = NULL) +
  theme(
    legend.title = element_text(colour = "grey30", size = rel(0.85)),
    legend.position = "top",
    panel.spacing = unit(1.1, "lines")
  )
```

## Season {.flow}

### {width="50%"}

```{r}
#| title: "Casual riders pile into summer; members ride all year"
#| fill: false
#| fig-width: 7.4
#| fig-height: 4.1
season_profile <- programme_month |>
  mutate(month = factor(
    month_num(year_month),
    levels = 1:12,
    labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
  )) |>
  group_by(programme, month) |>
  summarise(n_trips = sum(n_trips), .groups = "drop_last") |>
  mutate(share = n_trips / sum(n_trips)) |>
  ungroup()

ggplot(season_profile, aes(month, share, colour = programme, group = programme)) +
  annotate("rect", xmin = 4.5, xmax = 9.5, ymin = -Inf, ymax = Inf, fill = "grey88", alpha = 0.55) +
  geom_line(linewidth = 1.05) +
  geom_point(size = 2.2) +
  scale_colour_manual(values = pal) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
  labs(
    x = NULL, y = "Share of the group's trips",
    caption = "Shaded band is May–Sep. Months are pooled across 2021–2025; each line adds to 100%."
  )
```

### {width="50%"}

```{r}
#| title: "Winter trips are mostly members; summer trips split more evenly"
#| fill: false
#| fig-width: 7.4
#| fig-height: 4.1
month_mix <- programme_month |>
  mutate(month = factor(
    month_num(year_month),
    levels = 1:12,
    labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
  )) |>
  group_by(month, programme) |>
  summarise(n_trips = sum(n_trips), .groups = "drop_last") |>
  mutate(share = n_trips / sum(n_trips)) |>
  ungroup()

ggplot(month_mix, aes(month, share, fill = programme)) +
  geom_col(width = 0.84) +
  scale_fill_manual(values = pal) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02)),
    limits = c(0, 1)
  ) +
  labs(
    x = NULL, y = "Share of that month's trips",
    caption = "Who took the trips in each calendar month. Share of trips, not of people."
  )
```

## Mix {.flow}

### Bike type does not separate the two groups

```{r}
#| title: "Bike type does not separate the two groups"
#| fill: false
#| fig-width: 10
#| fig-height: 3.2
bike_mix <- bike_type |>
  mutate(
    programme = as_programme(member_casual),
    bike = recode(
      rideable_type,
      electric_bike = "Electric bike",
      classic_bike = "Classic bike",
      docked_bike = "Docked bike",
      electric_scooter = "Electric scooter",
      .default = tools::toTitleCase(gsub("_", " ", rideable_type))
    )
  ) |>
  group_by(programme, bike) |>
  summarise(n_trips = sum(n_trips), .groups = "drop_last") |>
  mutate(share = n_trips / sum(n_trips)) |>
  ungroup()

ggplot(bike_mix, aes(share, reorder(bike, share), fill = programme)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.62) +
  geom_text(
    aes(label = percent(share, accuracy = 1)),
    position = position_dodge(width = 0.72),
    hjust = -0.12, size = 3.4, colour = "grey20"
  ) +
  scale_fill_manual(values = pal) +
  scale_x_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 0.62),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    x = "Share of the group's trips", y = NULL,
    caption = "Classic and electric split almost evenly. Docked bikes are a casual-only leftover."
  )
```

# Data quality

## Quality {.flow}

### `r percent(drop_rate, accuracy = 0.1)` of `r comma(n_in)` source records were excluded {width="55%"}

```{r}
#| title: "2.3% of source records were excluded against a documented reason"
#| fill: false
#| fig-width: 8.4
#| fig-height: 4.3
ggplot(quality_month, aes(date, pct_kept)) +
  geom_line(colour = "#215C8C", linewidth = 1.05) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0.9, 1)) +
  date_axis +
  labs(x = NULL, y = "Share of source records kept")
```

### Sub-minute rides (dock glitches) account for most exclusions {width="45%"}

```{r}
#| title: "Sub-minute rides (dock glitches) account for most exclusions"
#| fill: false
#| fig-width: 7.2
#| fig-height: 4.3
reasons <- quality_month |>
  summarise(across(starts_with("n_") & !c(n_in, n_kept, n_dropped), sum)) |>
  pivot_longer(everything(), names_to = "reason", values_to = "n") |>
  filter(!grepl("station_kept$", reason), n > 0) |>
  mutate(
    reason = recode(
      reason,
      n_missing_ride_id = "Missing ride id",
      n_duplicate_ride_id = "Duplicate ride id",
      n_missing_programme = "Missing member/casual flag",
      n_missing_timestamps = "Unparseable timestamp",
      n_duration_non_positive = "End before start",
      n_duration_too_short = "Under 60 seconds",
      n_duration_too_long = "Over 24 hours"
    ),
    inside = n > max(n) * 0.45
  )

ggplot(reasons, aes(n, reorder(reason, n))) +
  geom_col(fill = "#8C6D46", width = 0.65) +
  geom_text(
    aes(
      label = comma(n),
      hjust = ifelse(inside, 1.12, -0.12),
      colour = ifelse(inside, "white", "grey25")
    ),
    size = 3.6
  ) +
  scale_colour_identity() +
  scale_x_continuous(
    labels = label_number(scale = 1e-3, suffix = "k"),
    expand = expansion(mult = c(0, 0.16))
  ) +
  labs(x = "Records excluded", y = NULL)
```

## Scope {.flow}

::: {.scope-note}
**Answered here.** How member and casual *trips* differ in timing, length, day of week, season and bike choice, across `r nrow(member_share)` months and `r comma(n_trips)` cleaned trips.

**Not answerable from this data.** Anything about *people*. Divvy's public files carry no rider identifier, so a rider taking 200 trips and 200 riders taking one trip each are indistinguishable. Every figure here is a share of trips, never a share of riders, and no conversion rate or per-rider frequency can be derived.

**Statistical note.** The member/casual difference in weekend riding is overwhelmingly significant on a chi-square test, but that test assumes independent observations and trips cluster within unobserved riders. The p-value is reported in the repository as descriptive only; the effect sizes on the first page are the meaningful result.

Full indicator definitions, exclusion rules and pipeline architecture are in [METHODS.md](https://github.com/ThacienneUwimanayantumye/Cyclistic-Bike-Share-Analysis/blob/main/docs/METHODS.md).
:::
