---
title: "Cyclistic Programme Monitor"
subtitle: "How annual members and casual riders use the bikes differently · Chicago Divvy · 2021–2025"
format:
  dashboard:
    theme: flatly
    orientation: rows
    embed-resources: true
    nav-buttons:
      - icon: journal-text
        href: https://github.com/ThacienneUwimanayantumye/Cyclistic-Bike-Share-Analysis/blob/main/docs/METHODS.md
        aria-label: Methods
      - icon: github
        href: https://github.com/ThacienneUwimanayantumye/Cyclistic-Bike-Share-Analysis
        aria-label: Source
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(ggplot2)
  library(tidyr)
  library(scales)
})

derived <- file.path("..", "data", "derived")
if (!file.exists(file.path(derived, "member_share.csv"))) {
  derived <- file.path("data", "derived")
}

read_derived <- function(name) {
  read_csv(file.path(derived, paste0(name, ".csv")), show_col_types = FALSE)
}

member_share <- read_derived("member_share")
programme_month <- read_derived("programme_month")
quality_month <- read_derived("quality_month")
timing <- read_derived("timing")
duration_month <- read_derived("duration_month")
duration_bins <- read_derived("duration_bins")
bike_type <- read_derived("bike_type")
rider_profile <- read_derived("rider_profile")

pal <- c(Member = "#215C8C", Casual = "#C05621")

theme_set(
  theme_minimal(base_size = 14) +
    theme(
      legend.position = "top",
      legend.title = element_blank(),
      legend.margin = margin(0, 0, 0, 0),
      panel.grid.minor = element_blank(),
      plot.subtitle = element_text(colour = "grey30", margin = margin(b = 8)),
      plot.caption = element_text(colour = "grey45", hjust = 0),
      axis.title = element_text(colour = "grey30", size = rel(0.9))
    )
)

#' Capitalised programme labels read better in legends than the raw codes.
as_programme <- function(x) factor(ifelse(x == "member", "Member", "Casual"),
                                   levels = c("Member", "Casual"))

to_date <- function(ym) as.Date(paste0(ym, "-01"))

for (nm in c("member_share", "programme_month", "duration_month", "quality_month")) {
  assign(nm, mutate(get(nm), date = to_date(year_month)))
}
programme_month$programme <- as_programme(programme_month$member_casual)
duration_month$programme <- as_programme(duration_month$member_casual)

#' Keep the x-axis inside the observed period; the default expansion
#' pushed an empty 2026 tick onto every time series.
date_axis <- scale_x_date(
  date_breaks = "1 year", date_labels = "%Y",
  expand = expansion(mult = c(0.01, 0.01))
)

#' Subtitles are clipped rather than wrapped inside a dashboard card.
wrap_sub <- function(x, width = 62) paste(strwrap(x, width), collapse = "\n")

n_trips <- sum(member_share$n_trips)
n_in <- sum(quality_month$n_in)
drop_rate <- 1 - n_trips / n_in

#' Share of each group's trips meeting a condition, as a named vector.
share_by_programme <- function(data, condition) {
  data |>
    mutate(programme = as_programme(member_casual)) |>
    group_by(programme) |>
    summarise(value = sum(n_trips[{{ condition }}]) / sum(n_trips), .groups = "drop") |>
    tibble::deframe()
}

weekend <- share_by_programme(timing, weekday %in% c("Saturday", "Sunday"))
peak <- share_by_programme(
  rider_profile,
  day_type == "Weekday" & time_block %in% c("AM peak", "PM peak")
)
long_trip <- share_by_programme(duration_bins, !duration_bin %in% c("1-5", "5-10", "10-15", "15-20", "20-30"))

med_duration <- duration_month |>
  group_by(programme) |>
  summarise(med = weighted.mean(duration_median_min, n_trips)) |>
  tibble::deframe()

commuter_like <- rider_profile |>
  filter(
    day_type == "Weekday",
    time_block %in% c("AM peak", "PM peak"),
    duration_band == "Under 15 min"
  ) |>
  group_by(member_casual) |>
  summarise(n = sum(n_trips), .groups = "drop")
casual_commuter_n <- commuter_like$n[commuter_like$member_casual == "casual"]
```

# The difference

## Headline {height="20%"}

```{r}
#| content: valuebox
#| title: "Casual trips taken at a weekend"
list(icon = "sun", color = "#C05621", value = percent(weekend[["Casual"]], accuracy = 0.1))
```

```{r}
#| content: valuebox
#| title: "Member trips taken at a weekend"
list(icon = "briefcase", color = "#215C8C", value = percent(weekend[["Member"]], accuracy = 0.1))
```

```{r}
#| content: valuebox
#| title: "Median casual trip"
list(icon = "hourglass-split", color = "#E8C39E",
     value = paste0(number(med_duration[["Casual"]], accuracy = 0.1), " min"))
```

```{r}
#| content: valuebox
#| title: "Median member trip"
list(icon = "hourglass", color = "#AFC6DA",
     value = paste0(number(med_duration[["Member"]], accuracy = 0.1), " min"))
```

## Signature {height="44%"}

### Weekday rides through the day {width="55%"}

```{r}
hourly <- timing |>
  filter(!weekday %in% c("Saturday", "Sunday")) |>
  mutate(programme = as_programme(member_casual)) |>
  group_by(programme, hour) |>
  summarise(n_trips = sum(n_trips), .groups = "drop_last") |>
  mutate(share = n_trips / sum(n_trips)) |>
  ungroup()

ggplot(hourly, aes(hour, share, colour = programme)) +
  annotate("rect", xmin = 6, xmax = 9, ymin = -Inf, ymax = Inf, fill = "grey88", alpha = 0.5) +
  annotate("rect", xmin = 16, xmax = 19, ymin = -Inf, ymax = Inf, fill = "grey88", alpha = 0.5) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = pal) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = seq(0, 23, 3), labels = function(h) sprintf("%02d:00", h)) +
  labs(
    subtitle = wrap_sub("Members peak twice, at the morning and the evening commute. Casual riders build to a single afternoon hump."),
    x = "Hour the trip started", y = "Share of the group's weekday trips",
    caption = "Shaded bands are the 06–09 and 16–19 commute windows."
  )
```

### Where the two groups diverge {width="45%"}

```{r}
signature <- tibble::tibble(
  metric = factor(
    rep(c("Ridden at a weekend", "Ridden in a weekday\ncommute window", "Longer than 30 minutes"), each = 2),
    levels = c("Longer than 30 minutes", "Ridden in a weekday\ncommute window", "Ridden at a weekend")
  ),
  programme = factor(rep(c("Member", "Casual"), 3), levels = c("Member", "Casual")),
  value = c(weekend[["Member"]], weekend[["Casual"]],
            peak[["Member"]], peak[["Casual"]],
            long_trip[["Member"]], long_trip[["Casual"]])
)

ggplot(signature, aes(value, metric, fill = programme)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.62) +
  geom_text(
    aes(label = percent(value, accuracy = 1)),
    position = position_dodge(width = 0.7), hjust = -0.2, size = 4, colour = "grey25"
  ) +
  scale_fill_manual(values = pal) +
  scale_x_continuous(labels = percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.18))) +
  labs(
    subtitle = wrap_sub("Casual riding is weekend, leisurely and long. Member riding is weekday, scheduled and short."),
    x = "Share of the group's trips", y = NULL
  )
```

## Opportunity {height="36%"}

### Casual trips that already follow a commute pattern {width="55%"}

```{r}
commute_trend <- rider_profile |>
  mutate(programme = as_programme(member_casual)) |>
  group_by(programme, year_month) |>
  summarise(
    share = sum(n_trips[day_type == "Weekday" &
                          time_block %in% c("AM peak", "PM peak") &
                          duration_band == "Under 15 min"]) / sum(n_trips),
    .groups = "drop"
  ) |>
  mutate(date = to_date(year_month))

ggplot(commute_trend, aes(date, share, colour = programme)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = pal) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
  date_axis +
  labs(
    subtitle = wrap_sub(paste0(
      comma(casual_commuter_n, scale = 1e-6, accuracy = 0.1, suffix = "M"),
      " casual trips over five years are short, weekday, peak-hour rides — the conversion target."
    )),
    x = NULL, y = "Share of the group's trips",
    caption = "Commute pattern = weekday, started 06–09 or 16–19, under 15 minutes."
  )
```

### How long a trip lasts {width="45%"}

```{r}
dur_dist <- duration_bins |>
  mutate(programme = as_programme(member_casual)) |>
  group_by(programme, duration_bin) |>
  summarise(n_trips = sum(n_trips), .groups = "drop_last") |>
  mutate(share = n_trips / sum(n_trips)) |>
  ungroup() |>
  mutate(duration_bin = factor(
    duration_bin,
    levels = c("1-5", "5-10", "10-15", "15-20", "20-30", "30-45", "45-60", "60-90", "90-120", "120+")
  ))

ggplot(dur_dist, aes(duration_bin, share, fill = programme)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  scale_fill_manual(values = pal) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    subtitle = wrap_sub("Member trips bunch under 15 minutes; casual trips carry a long tail."),
    x = "Trip length (minutes)", y = "Share of the group's trips"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Week and season

## Heatmaps {height="52%"}

### When each group rides, hour by hour

```{r}
heat <- timing |>
  mutate(programme = as_programme(member_casual)) |>
  group_by(programme, weekday, hour) |>
  summarise(n_trips = sum(n_trips), .groups = "drop_last") |>
  mutate(share = n_trips / sum(n_trips)) |>
  ungroup() |>
  mutate(weekday = factor(weekday, levels = rev(c(
    "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"
  ))))

ggplot(heat, aes(hour, weekday, fill = share)) +
  geom_tile() +
  facet_wrap(~programme) +
  scale_fill_viridis_c(
    option = "mako", direction = -1, labels = percent_format(accuracy = 0.1),
    name = "Share of the group's trips",
    guide = guide_colourbar(barwidth = 12, barheight = 0.6, title.position = "top")
  ) +
  scale_x_continuous(breaks = seq(0, 21, 3), labels = function(h) sprintf("%02d", h),
                     expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    subtitle = wrap_sub("Both panels share one scale, so the two patterns are directly comparable."),
    x = "Hour the trip started", y = NULL
  ) +
  theme(legend.title = element_text(colour = "grey30", size = rel(0.85)))
```

## Seasonality {height="48%"}

### Trips per month

```{r}
ggplot(programme_month, aes(date, n_trips, colour = programme)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = pal) +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  date_axis +
  labs(
    subtitle = wrap_sub("Casual volume collapses every winter; member volume holds up."),
    x = NULL, y = "Trips per month"
  )
```

### Member share of trips

```{r}
ggplot(member_share, aes(date, member_share)) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), fill = "#215C8C", alpha = 0.25) +
  geom_line(colour = "#215C8C", linewidth = 1) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  date_axis +
  labs(
    subtitle = wrap_sub("Shaded band is the Wilson 95% interval. Members dominate winter; casual riders close the gap each summer."),
    x = NULL, y = "Member share of trips",
    caption = "Share of trips, not of unique people — Divvy files carry no rider identifier."
  )
```

### Bike type chosen

```{r}
bike_mix <- bike_type |>
  mutate(programme = as_programme(member_casual),
         bike = tools::toTitleCase(gsub("_", " ", rideable_type))) |>
  group_by(programme, bike) |>
  summarise(n_trips = sum(n_trips), .groups = "drop_last") |>
  mutate(share = n_trips / sum(n_trips)) |>
  ungroup()

ggplot(bike_mix, aes(share, reorder(bike, share), fill = programme)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  scale_fill_manual(values = pal) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    subtitle = wrap_sub("Classic and electric split almost evenly for both groups. Docked bikes are a casual-only legacy option."),
    x = "Share of the group's trips", y = NULL
  )
```

# Data quality

## Quality {height="55%"}

### Records kept after the cleaning protocol

```{r}
ggplot(quality_month, aes(date, pct_kept)) +
  geom_line(colour = "#215C8C", linewidth = 1) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0.9, 1)) +
  date_axis +
  labs(
    subtitle = wrap_sub(paste0(
      percent(drop_rate, accuracy = 0.1), " of the ", comma(n_in),
      " source records were excluded, every one of them counted against a documented reason."
    )),
    x = NULL, y = "Share of source records kept"
  )
```

### Why records were excluded

```{r}
reasons <- quality_month |>
  summarise(across(starts_with("n_") & !c(n_in, n_kept, n_dropped), sum)) |>
  pivot_longer(everything(), names_to = "reason", values_to = "n") |>
  filter(!grepl("station_kept$", reason), n > 0) |>
  mutate(reason = recode(
    reason,
    n_missing_ride_id = "Missing ride id",
    n_duplicate_ride_id = "Duplicate ride id",
    n_missing_programme = "Missing member/casual flag",
    n_missing_timestamps = "Unparseable timestamp",
    n_duration_non_positive = "End before start",
    n_duration_too_short = "Under 60 seconds",
    n_duration_too_long = "Over 24 hours"
  ))

ggplot(reasons, aes(n, reorder(reason, n))) +
  geom_col(fill = "#8C6D46", width = 0.65) +
  geom_text(aes(label = comma(n)), hjust = -0.15, size = 4, colour = "grey25") +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = "k"),
                     expand = expansion(mult = c(0, 0.2))) +
  labs(
    subtitle = wrap_sub("Sub-minute rides, meaning dock glitches and false starts, account for most exclusions."),
    x = "Records excluded", y = NULL
  )
```

## Scope {height="45%"}

### What this dashboard can and cannot tell you

**Answered here.** How member and casual *trips* differ in timing, length, day of week, season and bike choice, across `r nrow(member_share)` months and `r comma(n_trips)` cleaned trips.

**Not answerable from this data.** Anything about *people*. Divvy's public files carry no rider identifier, so a rider taking 200 trips and 200 riders taking one trip each are indistinguishable. Every figure here is a share of trips, never a share of riders, and no conversion rate or per-rider frequency can be derived.

**Statistical note.** The member/casual difference in weekend riding is overwhelmingly significant on a chi-square test, but that test assumes independent observations and trips cluster within unobserved riders. The p-value is reported in the repository as descriptive only; the effect sizes above are the meaningful result.

Full indicator definitions, exclusion rules and pipeline architecture are in [METHODS.md](https://github.com/ThacienneUwimanayantumye/Cyclistic-Bike-Share-Analysis/blob/main/docs/METHODS.md).
