---
title: "Cyclistic Programme Monitor"
subtitle: "Methods and findings, 2021–2025"
author: "Portfolio data product"
date: today
format:
  html:
    toc: true
    theme: flatly
    code-fold: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(ggplot2)
  library(tidyr)
  library(here)
})
source(here::here("R", "config.R"))
source(here::here("dashboard", "helpers.R"))
# helpers.R derived_dir() uses here() — derived tables must exist
member_share <- read_indicator("member_share")
programme_month <- read_indicator("programme_month")
quality_month <- read_indicator("quality_month")
timing <- read_indicator("timing")
duration_month <- read_indicator("duration_month")
bike_type <- read_indicator("bike_type")
volume_stl <- read_indicator("volume_stl")
chisq_weekend <- read_indicator("chisq_weekend")
```

# Purpose

Cyclistic’s finance team treats **annual members** as more profitable than **casual** riders (single-ride and day passes). A monitoring product should show **how the two programmes are used**, with uncertainty and data quality, not only a pair of bar charts.

This report uses five calendar years of public Divvy trip files (January 2021–December 2025). The unit is the **trip**. There is no rider identifier, so we do not estimate unique participants or “coverage.”

The original student analysis (2019–2020 quarterly files, mean ride length, two plots) is in `legacy/`.

# Methods

Cleaning rules, indicator definitions, Wilson intervals, STL, and the chi-square caveat are specified in [`docs/METHODS.md`](../docs/METHODS.md). Headline duration is the **median**, after dropping trips shorter than 60 seconds or longer than 24 hours.

Monthly files are processed independently; only indicator tables are combined. The dashboard reads those tables, not microdata.

# Scale and quality

```{r}
#| label: scale
overall <- member_share |>
  summarise(
    n_trips = sum(n_trips),
    n_member = sum(n_member),
    n_casual = sum(n_casual),
    member_share = n_member / n_trips
  )
quality <- quality_month |>
  summarise(
    n_in = sum(n_in),
    n_kept = sum(n_kept),
    n_dropped = sum(n_dropped),
    drop_rate = n_dropped / n_in
  )
knitr::kable(
  tibble::tibble(
    Metric = c("Source records", "Kept trips", "Drop rate", "Member trip share"),
    Value = c(
      scales::comma(quality$n_in),
      scales::comma(quality$n_kept),
      scales::percent(quality$drop_rate, accuracy = 0.01),
      scales::percent(overall$member_share, accuracy = 0.1)
    )
  ),
  caption = "Analytic extract, 2021–2025"
)
```

```{r}
#| label: fig-quality
#| fig-cap: "Monthly drop rate (hard exclusions) and missing start station among kept trips."
q <- quality_month |>
  mutate(drop_rate = n_dropped / n_in) |>
  pivot_longer(c(drop_rate, pct_missing_start_station), names_to = "series", values_to = "value")
ggplot(q, aes(year_month, value, group = series, colour = series)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  scale_colour_manual(
    values = c(drop_rate = "#7B2D26", pct_missing_start_station = "#215C8C"),
    labels = c(drop_rate = "Drop rate", pct_missing_start_station = "Missing start station (kept)"),
    name = NULL
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 6), legend.position = "top")
```

Missing start stations among kept trips are expected for many electric / dockless events; they are a quality flag, not a reason to drop the ride.

# Programme mix over time

```{r}
#| label: fig-share
#| fig-cap: "Member share of trips, with Wilson 95% intervals. Intervals are narrow because monthly n is large; they still document the method."
ggplot(member_share, aes(year_month, member_share, group = 1)) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), fill = "#215C8C", alpha = 0.2) +
  geom_line(colour = "#215C8C") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = NULL, y = "Member trip share") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 6))
```

```{r}
#| label: fig-volume
#| fig-cap: "Monthly trip volume by programme."
ggplot(programme_month, aes(year_month, n_trips, colour = member_casual, group = member_casual)) +
  geom_line() +
  scale_colour_manual(values = c(member = "#215C8C", casual = "#C05621"), name = NULL) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Trips") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 6), legend.position = "top")
```

```{r}
#| label: fig-stl
#| fig-cap: "STL decomposition of total monthly trips (trend vs seasonal vs remainder)."
stl_long <- volume_stl |>
  pivot_longer(c(observed, trend, seasonal, remainder), names_to = "component", values_to = "value")
ggplot(stl_long, aes(year_month, value, group = 1)) +
  geom_line(colour = "#215C8C") +
  facet_wrap(~component, ncol = 1, scales = "free_y") +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 6))
```

Volume is strongly seasonal. Any “growth” claim must look at the **trend** component, not a single summer month.

# When trips start

```{r}
#| label: fig-heat
#| fig-cap: "Weekday by hour of trip start, summed over 2021–2025. Members concentrate on weekday daytime and evening peaks; casual use is heavier on weekends."
heat <- timing |>
  group_by(member_casual, weekday, hour) |>
  summarise(n_trips = sum(n_trips), .groups = "drop") |>
  mutate(weekday = factor(weekday, levels = WEEKDAY_LEVELS))
ggplot(heat, aes(hour, weekday, fill = n_trips)) +
  geom_tile() +
  facet_wrap(~member_casual, ncol = 1) +
  scale_fill_gradient(low = "#F4F7FA", high = "#215C8C") +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  labs(x = "Hour", y = NULL, fill = "Trips") +
  theme_minimal()
```

```{r}
#| label: chisq
x <- chisq_weekend[1, ]
knitr::kable(
  tibble::tibble(
    Item = c("Chi-square", "df", "p-value", "Note"),
    Value = c(
      as.character(round(x$statistic, 1)),
      as.character(x$df),
      format.pval(x$p_value, eps = 1e-16),
      x$note
    )
  ),
  caption = "Programme × weekend, trip counts (descriptive only)"
)
```

This finally measures the commute-vs-leisure pattern that the original report asserted without an hour-of-day table.

# Duration

```{r}
#| label: fig-duration
#| fig-cap: "Median trip duration (minutes) by month and programme. Casual trips remain longer; the gap is smaller than the original mean-based comparison because extremes are capped and the median is used."
ggplot(duration_month, aes(year_month, duration_median_min, colour = member_casual, group = member_casual)) +
  geom_line() +
  scale_colour_manual(values = c(member = "#215C8C", casual = "#C05621"), name = NULL) +
  labs(x = NULL, y = "Median minutes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 6), legend.position = "top")
```

# Bike type

```{r}
#| label: fig-bikes
#| fig-cap: "Trip counts by bicycle type and programme (all years combined)."
b <- bike_type |>
  group_by(member_casual, rideable_type) |>
  summarise(n_trips = sum(n_trips), .groups = "drop")
ggplot(b, aes(rideable_type, n_trips, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(member = "#215C8C", casual = "#C05621"), name = NULL) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Trips") +
  theme_minimal() +
  theme(legend.position = "top")
```

# What this would mean for a membership team

These are **monitoring conclusions**, not a campaign brief:

1. Casual use is more weekend- and midday-shaped; member use is more weekday-shaped. That is now visible in the hour heatmap, not only in weekday totals.
2. Casual trips are longer on the median, but the original analysis overstated the gap by using the mean of uncapped durations.
3. Electric bikes are a large share of events; station-based metrics miss GPS-only trips if missing station names are dropped.

A registry-style next step (not possible here) would be a **person key** and a true participation rate.

# Reproducibility

`targets::tar_make()` rebuilds DuckDB and `data/derived/`. `shiny::runApp("dashboard")` is the interactive product. Tests live in `tests/testthat/`. Agentic-AI use is recorded in `AI.md`.
